Decompositions of congruence subgroups \\ of Chevalley groups

### 2018-01-30 paper.tex
Invented simpler proof in the case F_4

\section{Introduction}\label{sec:intro}
The aim of this paper is to generalize several decompositions of Chevalley groups over rings 
 to their principal congruence subgroups and relative elementary subgroups.
We focus on the so-called ``parabolic decompositions'', i.e. decompositions involving parabolic and unipotent factors. 
One of our main results is the following theorem.
\begin{thm}[Dennis--Vaserstein decomposition]\label{thm:DennisVaserstein}
Let $\Phi$ be a either a classical irreducible root system of rank $\ell\geqslant 2$ or $\Phi=\rE_7, \rE_8$. 
Let $\{ r, s \}$ be a pair of distinct endnodes of the Dynkin diagram of $\Phi$.
Denote the distance between $r$ and $s$ by $d$.
Assume that $\Phi$, $I$ and $\{r, s\}$ satisfy either of the following assumptions:
\begin{thmlist}
 \item $\sr(I) \leqslant d$ for classical $\Phi$,
 \item $\sr(I) \leqslant d$ for $\Phi=\rE_\ell$ with $\{r, s\} = \{ 2, \ell \}$,
 \item $\asr(I) \leqslant d$ for $\Phi=\rE_\ell$ with $\{r, s\} = \{1, \ell \}$,
\end{thmlist}
Then $\E(\Phi, R, I) = \EP_r(R, I) \cdot \U(\Sigma^-_r \cap \Sigma^-_s, I) \cdot \EP_s(R, I).$
\end{thm}
Even in the special case $I=R$ the above theorem contains several new cases as compared to~\cite[Theorems~2.5 and~4.1]{St78}.
The condition $\asr(I) \leq d$ is a relative version of the \emph{absolute stable range} condition of D.~Estes and J.~Ohm, which depends only on the ideal $I$ (see section~\ref{sec:rel-asr}).
Another results of the paper are relative analogues of Gauss and Bass--Kolster decompositions (see~\cref{thm:srRI1,thm:BassKolster}).

@@

\begin{proof} [Proof of \cref{thm:DennisVaserstein}]
Since the equality $A_{rs} = \E(\Phi, R, I)$ implies $A_{sr} = \E(\Phi, R, I)$, it suffices to consider only the possibilities for $\Phi$, $s$, $r$ listed in~\cref{table:dv-reps}.
\begin{table}[htb]
\[\begin{array}{l @{\qquad} l @{\qquad} c @{\quad} c @{\quad} c @{\qquad} c @{\qquad} c @{\qquad} c}
\Phi                                 & s    &r      & |\Lambda(\pi)| & \text{type of $\pi$} & \text{type of $\Delta_r$} & |\Lambda(\pi')|& |\Gamma|  \\ \hline\vphantom{\Bigl(}
\rA_\ell, \ \ell\geqslant 2           & 1    &\ell   & \ell+1         & \text{natural}       & \rA_{\ell-1}              & \ell           & 1  \\     
\rB_\ell, \ \ell\geqslant 2           & 1    &\ell   & 2\ell+1        & \text{natural}       & \rA_{\ell-1}              & \ell           & 1  \\     
\rC_\ell, \ \ell\geqslant 2           & 1    &\ell   & 2\ell          & \text{natural}       & \rA_{\ell-1}              & \ell           & 1  \\
\rD_\ell, \ \ell\geqslant 4           & 1    &\ell   & 2\ell          & \text{natural}       & \rA_{\ell-1}              & \ell           & 2  \\ 
\rD_\ell, \ \ell\geqslant 4           & \ell &\ell-1 & 2^{\ell-1}     & \text{half-spinor}   & \rA_{\ell-1}              & \ell           & \ell-2  \\
\rE_\ell, \ \ell=6, 7                  & \ell &2      & 27, 56         & \text{minimal}       & \rA_{\ell-1}              & \ell           & 3       \\ 
\rE_\ell, \ \ell=6, 7                  & \ell &1      & 27, 56         & \text{minimal}       & \rD_{\ell-1}              & 2(\ell-1)      & \ell-1  \end{array}\]
 \caption{List of representations used in the proof of \cref{thm:DennisVaserstein}.} \label{table:dv-reps}
\end{table}

@@

First of all notice that $N_{rs}$ is sufficiently large.
\cref{lemma:Stein_reduction} below guarantees that $N_{rs}$ contains the root subgroup $X_{-\alpha_{r}}(R)$, which together with~\cref{lemma:dv-normal} 
 implies that $\EP_s(R) \subseteq N_{rs}$ and consequently that $\EP_s(R, I) \subseteq L_{rs}$. Indeed for $g\in \EP_s(R, I) \subseteq \EP_s(R)$ one has
$g \cdot L_{rs} = g \cdot L_{rs} \cdot g^{-1} \cdot g \subseteq L_{rs} \cdot g \subseteq L_{rs}.$

@@

The next step of the proof is to show that $\U(\Sigma_s^-, I) \subseteq L_{rs}$.
All the possibilities for $\Phi$, $s$ and $r$ fall into one of the following three cases:
\begin{enumerate}
 \item \textit{Case $m_s(\alpha_{\mathrm{max}})=1$ and $\Phi$ is simply laced.}
  In this case all the roots of $\Sigma^-_s$ have the same $s$-shape and length.
  Since $\widetilde{W}(\Delta_s) \subseteq \E(\Delta_s, R) \subseteq N_{rs}$ and $X_{-\alpha_s}(I) \subseteq \EP_r(R, I) \subseteq L_{rs}$,
   the required inclusion $\U(\Sigma_s^-, I) \subseteq L_{rs}$ follows from \cref{item-trans2}.  
 \item \textit{Case $\Phi = \rB_\ell$, $s=1$, $r=\ell$.} 
  The argument from the previous case shows that $L_{\ell,1}$ contains root subgroups $X_\alpha(I)$, where $\alpha$ varies over the long roots of $\Sigma_1^-$. 
  The inclusion $X_{0, 1}(I) \subseteq L_{\ell, 1}$ for the remaining short root follows from inclusions $X_{-1, 2}(I) \cup X_{2, 1}(I) \subseteq L_{\ell,1}$, $X_{2, 0}(R) \subseteq N_{\ell,1}$
  using Chevalley commutator formula $x_{0, 1}(-ab) = [x_{-1,2}(a),\, x_{2,0}(b)] \cdot x_{2,1}(ab^2).$
 \item \textit{Case $\Phi = \rC_\ell$, $s=1$, $r=\ell$.} 
  Similarly, one can show that $L_{\ell,1}$ contains $X_\alpha(I)$ for all short $\alpha \in \Sigma_1^-$.
  However, the inclusion $X_{-1, 1}(I) \subseteq L_{\ell, 1}$ for the remaining long root does not immediately follow from commutator formulae.
  To verify it we invoke stability conditions once again.
  From~\cref{lemma:DVST} we obtain that $N_{\ell, 1}$ contains $X_{-\alpha\ssub{1}}(R)$ and therefore coincides with the whole $\E(\Phi, R)$, which gives the required assertion.
\end{enumerate}

Thus, we have just shown that $\mathcal{X}\subseteq L_{rs}$. 
Recall from~\cite[Theorem~3.4]{S} that the group $\E(\Phi, R, I)$ is generated by $\mathcal{Z}(\Sigma_s^-)$.
On the other hand, from $\mathcal{Z}(\Sigma_s^-) \subseteq \mathcal{X}^{\EP_s(R)} \subseteq L_{rs}^{N_{rs}} \subseteq L_{rs}$ we conclude that 
 $\E(\Phi, R, I) \subseteq L_{rs}$, which implies the assertion of the theorem.
%Thus, to prove our assertion it suffices to show that $L_{rs}$ contains $\mathcal{Z}(\Sigma_s^-)$.
%From the inclusion  and \eqref{rel:NnormL} it follows that we only need to verify the inclusion $\mathcal{X} \subseteq L_{rs}$.
%In turn, the last inclusion follows from the same transitivity argument as above if we use the fact that $X_\alpha(I) \subseteq \EP_r(R, I) \subseteq L_{rs}$ for $\alpha \in \Delta_r \cup \Sigma_r$.
%For $\Phi=\rB_\ell$ we need to apply Chevalley commutator formula once again:
%\[ X_{0, 1}(I) \subseteq X_{2, 1}(I) \cdot X_{2, 1}(I)^{x_{0, 2}(1)} \cdot X_{-2, 1}(I) \subseteq L_{rs} \cdot L_{rs}^{N_{rs}} \cdot L_{rs} \subseteq L_{rs}. \qedhere \] 
\end{proof}

>>>>

\section{Introduction}\label{sec:intro}
The aim of this paper is to generalize several decompositions of Chevalley groups over rings 
 to their principal congruence subgroups and relative elementary subgroups.
We focus on the so-called ``parabolic decompositions'', i.e. decompositions involving parabolic and unipotent factors. 
One of our main results is the following theorem.
\begin{thm}[Dennis--Vaserstein decomposition]\label{thm:DennisVaserstein}
Let $\Phi$ be a either a classical irreducible root system of rank $\ell\geqslant 2$ or $\Phi=\rE_7, \rE_8$. 
Let $\{ r, s \}$ be a pair of distinct endnodes of the Dynkin diagram of $\Phi$.
Denote the distance between $r$ and $s$ by $d$.
Assume that $\Phi$, $I$ and $\{r, s\}$ satisfy either of the following assumptions:
\begin{thmlist}
 \item $\sr(I) \leqslant d$ for classical $\Phi$,
 \item $\sr(I) \leqslant d$ for $\Phi=\rE_\ell$ with $\{r, s\} = \{ 2, \ell \}$,
 \item $\asr(I) \leqslant d$ for $\Phi=\rE_\ell$ with $\{r, s\} = \{1, \ell \}$,
 \item $\sr(I) \leqslant d$ for $\Phi = \rF_4$ with $\{r,s\} = \{1, 4\}$.
\end{thmlist}
Then $\E(\Phi, R, I) = \EP_r(R, I) \cdot \U(\Sigma^-_r \cap \Sigma^-_s, I) \cdot \EP_s(R, I).$
\end{thm}
Even in the special case $I=R$ the above theorem contains several new cases as compared to~\cite[Theorems~2.5 and~4.1]{St78}.
The condition $\asr(I) \leq d$ is a relative version of the \emph{absolute stable range} condition of D.~Estes and J.~Ohm, which depends only on the ideal $I$ (see section~\ref{sec:rel-asr}).
Another results of the paper are relative analogues of Gauss and Bass--Kolster decompositions (see~\cref{thm:srRI1,thm:BassKolster}).

@@

\begin{proof} [Proof of \cref{thm:DennisVaserstein}]
Since the equality $A_{rs} = \E(\Phi, R, I)$ implies $A_{sr} = \E(\Phi, R, I)$, it suffices to consider only the possibilities for $\Phi$, $s$, $r$ listed in~\cref{table:dv-reps}.
\begin{table}[htb]
\[\begin{array}{l @{\qquad} l @{\qquad} c @{\quad} c @{\quad} c @{\qquad} c @{\qquad} c @{\qquad} c}
\Phi                                 & s    &r      & |\Lambda(\pi)| & \text{type of $\pi$} & \text{type of $\Delta_r$} & |\Lambda(\pi')|& |\Gamma|  \\ \hline\vphantom{\Bigl(}
\rA_\ell, \ \ell\geqslant 2           & 1    &\ell   & \ell+1         & \text{natural}       & \rA_{\ell-1}              & \ell           & 1  \\     
\rB_\ell, \ \ell\geqslant 2           & 1    &\ell   & 2\ell+1        & \text{natural}       & \rA_{\ell-1}              & \ell           & 1  \\     
\rC_\ell, \ \ell\geqslant 2           & 1    &\ell   & 2\ell          & \text{natural}       & \rA_{\ell-1}              & \ell           & 1  \\
\rD_\ell, \ \ell\geqslant 4           & 1    &\ell   & 2\ell          & \text{natural}       & \rA_{\ell-1}              & \ell           & 2  \\ 
\rD_\ell, \ \ell\geqslant 4           & \ell &\ell-1 & 2^{\ell-1}     & \text{half-spinor}   & \rA_{\ell-1}              & \ell           & \ell-2  \\
\rE_\ell, \ \ell=6, 7                  & \ell &2      & 27, 56         & \text{minimal}       & \rA_{\ell-1}              & \ell           & 3       \\ 
\rE_\ell, \ \ell=6, 7                  & \ell &1      & 27, 56         & \text{minimal}       & \rD_{\ell-1}              & 2(\ell-1)      & \ell-1 \\
\rF_4,                                 & 4    & 1     & 26              & \text{minimal}       & \rC_{3}                  & 6              & 3 \end{array}\]
 \caption{List of representations used in the proof of \cref{thm:DennisVaserstein}.} \label{table:dv-reps}
\end{table}

@@

\cref{lemma:Stein_reduction} below guarantees that $N_{rs}$ contains the root subgroup $X_{-\alpha_{r}}(R)$, which together with~\cref{lemma:dv-normal} 
 implies that $\EP_s(R) \subseteq N_{rs}$ and consequently that $\EP_s(R, I) \subseteq L_{rs}$. Indeed for $g\in \EP_s(R, I) \subseteq \EP_s(R)$ one has
$g \cdot L_{rs} = g \cdot L_{rs} \cdot g^{-1} \cdot g \subseteq L_{rs} \cdot g \subseteq L_{rs}.$

@@

The next step of the proof is to show that $\U(\Sigma_s^-, I) \subseteq L_{rs}$.
All the possibilities for $\Phi$, $s$ and $r$ fall into one of the following three cases:
\begin{enumerate}
 \item \textit{Case $m_s(\alpha_{\mathrm{max}})=1$ and $\Phi$ is simply laced.}
  In this case all the roots of $\Sigma^-_s$ have the same $s$-shape and length.
  Since $\widetilde{W}(\Delta_s) \subseteq \E(\Delta_s, R) \subseteq N_{rs}$ and $X_{-\alpha_s}(I) \subseteq \EP_r(R, I) \subseteq L_{rs}$,
   the required inclusion $\U(\Sigma_s^-, I) \subseteq L_{rs}$ follows from \cref{item-trans2}.  
 \item \textit{Case $\Phi = \rB_\ell$, $s=1$, $r=\ell$.} 
  The argument from the previous case shows that $L_{rs}$ contains root subgroups $X_\alpha(I)$, where $\alpha$ varies over the long roots of $\Sigma_1^-$.
  Consider the following two roots of $\Phi$:
  \begin{equation}\nonumber \beta = \alpha_2 + \ldots + \alpha_\ell,\ \gamma = -\alpha_1 - 2\alpha_2 - \ldots -2\alpha_\ell.\end{equation}
  We need to verify the inclusion $X_{\beta+\gamma}(I) \subseteq L_{rs}$ for the remaining short root of $\Sigma_1^-$.
  But the latter follows from $X_{\gamma}(I) \cup X_{-\alpha_1}(I) \subseteq L_{rs}$, $X_{\beta}(R) \subseteq N_{rs}$ using Chevalley commutator formula:
  \begin{equation} \label{eq:CCF} x_{\beta+\gamma}(-ab) = [x_{\gamma}(a),\, x_{\beta}(b)] \cdot x_{\gamma + 2\beta}(ab^2). \end{equation}
 \item \textit{Case $\Phi = \rF_4$, $s=4$, $r=1$.} In this case there are three possibilities for the $s$-shape and length of a root from $\Sigma_4^-$: 
 \begin{itemize}
  \item $\shape(s, \alpha) = 1$, $\alpha$ is short (there are 8 such roots);
  \item $\shape(s, \alpha) = 2$, $\alpha$ is long (there are 6 such roots);
  \item $\shape(s, \alpha) = 2$, $\alpha$ is short (there is only one such root: $\delta = -\alpha_1 -2\alpha_2-3\alpha_3-2\alpha_4$).
 \end{itemize}
 Since $\Delta_r \cap \Sigma_s^-$ contains roots of the first two types, we have the neccessary inclusions for the root subgroups of corresponding type via the same argument as above.
 To verify the inclusion for the remaining short root subgroup invoke~\eqref{eq:CCF} with $\beta = \alpha_2 + \alpha_3,\ \gamma = -\alpha_{\mathrm{max}}$. 
 \item \textit{Case $\Phi = \rC_\ell$, $s=1$, $r=\ell$.} 
  Similarly, one can show that $L_{rs}$ contains $X_\alpha(I)$ for all short $\alpha \in \Sigma_1^-$.
  However, the inclusion $X_{-1, 1}(I) \subseteq L_{rs}$ for the remaining long root does not immediately follow from commutator formulae and to verify it we need to invoke stability conditions once again.
  From~\cref{lemma:DVST} we obtain that $N_{rs}$ contains $X_{-\alpha_1}(R)$ and therefore coincides with $\E(\Phi, R)$, which gives the required assertion.
\end{enumerate}

Thus, we have just shown that $\mathcal{X}\subseteq L_{rs}$. 
Recall from~\cite[Theorem~3.4]{S} that the group $\E(\Phi, R, I)$ is generated by $\mathcal{Z}(\Sigma_s^-)$.
On the other hand, from $\mathcal{Z}(\Sigma_s^-) \subseteq \mathcal{X}^{\EP_s(R)} \subseteq L_{rs}^{N_{rs}} \subseteq L_{rs}$ we conclude that 
 $\E(\Phi, R, I) \subseteq L_{rs}$, which completes the proof.
%Thus, to prove our assertion it suffices to show that $L_{rs}$ contains $\mathcal{Z}(\Sigma_s^-)$.
%From the inclusion  and \eqref{rel:NnormL} it follows that we only need to verify the inclusion $\mathcal{X} \subseteq L_{rs}$.
%In turn, the last inclusion follows from the same transitivity argument as above if we use the fact that $X_\alpha(I) \subseteq \EP_r(R, I) \subseteq L_{rs}$ for $\alpha \in \Delta_r \cup \Sigma_r$.
%For $\Phi=\rB_\ell$ we need to apply Chevalley commutator formula once again:
%\[ X_{0, 1}(I) \subseteq X_{2, 1}(I) \cdot X_{2, 1}(I)^{x_{0, 2}(1)} \cdot X_{-2, 1}(I) \subseteq L_{rs} \cdot L_{rs}^{N_{rs}} \cdot L_{rs} \subseteq L_{rs}. \qedhere \] 
\end{proof}
